---
title: "Assignment 9: Spatial Analysis in R"
author: "Sydney Williams"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics
(ENV872L) on spatial analysis.

## Directions

1.  Rename this file `<FirstLast>_A09_SpatialAnalysis.Rmd` (replacing
    `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Use the lesson as a guide. It contains code that can be modified to
    complete the assignment.
4.  Work through the steps, **creating code and output** that fulfill
    each instruction.
5.  Be sure to **answer the questions** in this assignment document.
    Space for your answers is provided in this document and is indicated
    by the "\>" character. If you need a second paragraph be sure to
    start the first line with "\>". You should notice that the answer is
    highlighted in green by RStudio.
6.  When you have completed the assignment, **Knit** the text and code
    into a single HTML file.

## DATA WRANGLING

### Set up your session

1.  Import libraries: tidyverse, sf, leaflet, here, and mapview
2.  Execute the `here()` command to display the current project
    directory

```{r Prepare workspace, error=F, warning=F}
#1. 
library(tidyverse)
library(sf)
library(leaflet)
library(here)
library(mapview)

#2. 
here()

```

### Read (and filter) county features into an sf dataframe and plot

In this exercise, we will be exploring stream gage height data in
Nebraska corresponding to floods occurring there in 2019. First, we will
import from the US Counties shapefile we've used in lab lessons,
filtering it this time for just Nebraska counties. Nebraska's state FIPS
code is `31` (as North Carolina's was `37`).

3.  Read the `cb_2018_us_county_20m.shp` shapefile into an sf dataframe,
    filtering records for Nebraska counties (State FIPS = 31)
4.  Reveal the dataset's coordinate reference system
5.  Plot the records as a map (using `mapview` or `ggplot`)

```{r Read the county data into an sf dataframe}
#3. Read in Counties shapefile into an sf dataframe, filtering for just NE counties
NE_counties <- st_read(here('Data/Spatial/cb_2018_us_county_20m.shp')) %>%
    filter(STATEFP == "31")

#4. Reveal the CRS of the counties features
print(st_crs(NE_counties))

#5. Plot the data
NEcounties_plot <- ggplot() +
  geom_sf(data = NE_counties) +
  theme_minimal()

NEcounties_plot
```

6.  What is the EPSG code of the Counties dataset? Is this a geographic
    or a projected coordinate reference system? (Or, does this CRS use
    angular or planar coordinate units?) To what datum is this CRS
    associated? (Tip: look for the EPSG code on
    <https://spatialreference.org> and examine the link for "Well Known
    Text as HTML" or "Human-Readable OGC WKT"...)

> ANSWER: The EPSG code for the Counties dataset is 4269 and is a geographic coordinate reference system. The North American Datum 1983. 

### Read in gage locations csv as a dataframe, then display the column names it contains

Next we'll read in some USGS/NWIS gage location data added to the
`Data/Raw` folder. These are in the `NWIS_SiteInfo_NE_RAW.csv` file.(See
`NWIS_SiteInfo_NE_RAW.README.txt` for more info on this dataset.)

7.  Read the NWIS_SiteInfo_NE_RAW.csv file into a standard dataframe,
    being sure to set the `site_no` field as well as other character
    columns as a factor.

8.  Display the column names of this dataset.

```{r Read in USGS Gage Locations}
#7. Read in gage locations csv as a dataframe
gage_locationdata <- read.csv("Data/Raw/NWIS_SiteInfo_NE_RAW.csv", stringsAsFactors = TRUE)

#8. Reveal the names of the columns
str(gage_locationdata)

colnames(gage_locationdata)


```

9.  What columns in the dataset contain the x and y coordinate values,
    respectively?\
    \> ANSWER: \>dec_lat_va and dec_long_va. These are latitudinal and longitudinal points. 

### Convert the dataframe to a spatial features ("sf") dataframe

10. Convert the dataframe to an sf dataframe.

-   Note: These data use the same coordinate reference system as the
    counties dataset

11. Display the column names of the resulting sf dataframe

```{r Convert dataframe to sf dataframe}
#10. Convert to an sf object
gage_sf <- st_as_sf(gage_locationdata, coords = c("dec_long_va", "dec_lat_va"), crs = st_crs(NE_counties))

#11. Re-examine the column names
colnames(gage_sf)

```

12. What new field(s) appear in the sf dataframe created? What field(s),
    if any, disappeared?

> ANSWER: Geometry appears. dec_lat_va and dec_long_va disappears. 

### Plot the gage locations on top of the counties

13. Use `ggplot` to plot the county and gage location datasets.

-   Be sure the datasets are displayed in different colors
-   Title your plot "NWIS Gage Locations in Nebraska"
-   Subtitle your plot with your name

```{r Plot the spatial features}
#13. Plot the gage locations atop the county features
county_gagelocation_plot <- ggplot() +
  geom_sf(data = NE_counties, fill = "lightgreen") + 
  labs(title = "NWIS Gage Locations in Nebraska", subtitle = "By: [Sydney Williams]") +
  theme_minimal()

```

### Read in the gage height data and join the site location data to it.

Lastly, we want to attach some gage height data to our site locations.
I've constructed a csv file listing many of the Nebraska gage sites, by
station name and site number along with stream gage heights (in meters)
recorded during the recent flood event. This file is titled
`NWIS_SiteFlowData_NE_RAW.csv` and is found in the Data/Raw folder.

14. Read the `NWIS_SiteFlowData_NE_RAW.csv` dataset in as a dataframe

-   Pay attention to which fields should be imported as factors!

15. Show the column names .
16. Join our site information (already imported above) to these gage
    height data

-   The `site_no` and `station_nm` can both/either serve as joining
    attributes
-   Construct this join so that the result only includes spatial
    features where both tables have data

17. Show the column names in this resulting spatial features object
18. Show the dimensions of the resulting joined dataframe

```{r Read in gage site data}
#14. Read the site flow data into a data frame
siteflow_data <- read.csv("Data/Raw/NWIS_SiteFlowData_NE_RAW.csv", stringsAsFactors = TRUE)

#15. Show the column names
colnames(siteflow_data)

#16. Join the flow data to our NWIS gage location spatial dataframe
joined_flowdata <- merge(gage_locationdata, siteflow_data, by = "site_no", all.x = TRUE)

#17. Show the column names of the joined dataset
colnames(joined_flowdata)

#18. Show the dimensions of this joined dataset
dim(joined_flowdata)

```

### Map the pattern of gage height data

Now we can examine where the flooding appears most acute by visualizing
gage heights spatially. 19. Plot the gage sites on top of counties
(using `mapview`, `ggplot`, or `leaflet`) \* Show the magnitude of gage
height by color, shape, other visualization technique.

```{r map gage sites}
#Map the points, sized by gage height
gagesites_map <- ggplot() +
  geom_sf(data = NE_counties, fill = "lightpink") + 
  geom_point(data = joined_flowdata, aes(x = dec_long_va, y = dec_lat_va, color = gage_ht), size = 3) +  
  scale_color_viridis_c() + 
  labs(title = "Gage Sites with Gage Height in Nebraska") + 
  theme_minimal()  

gagesites_map

```

```

------------------------------------------------------------------------

## SPATIAL ANALYSIS

Up next we will do some spatial analysis with our data. To prepare for
this, we should transform our data into a projected coordinate system.
We'll choose UTM Zone 14N (EPGS = 32614).

### Transform the counties and gage site datasets to UTM Zone 14N

20. Transform the counties and gage sf datasets to UTM Zone 14N (EPGS =
    32614).
21. Using `mapview` or `ggplot`, plot the data so that each layer is
    shown with a unique color (e.g. counties blue and gages red)

```{r spatial analysis}
#20 Transform the counties and gage location datasets to UTM Zone 14

NECounties_sf_utm <- st_as_sf(NE_counties, coords = c("dec_long_va", "dec_lat_va"), crs = st_crs(NE_counties))

NEcounties_sf_utm <-st_transform(NE_counties,32614)
gage_sf_utm <-st_transform(gage_sf,32614)

#21 Plot the data
mapview(NEcounties_sf_utm, zcol = NULL, col.regions = "pink") + 
  mapview(gage_sf_utm, zcol = NULL, col.regions = "green")  

```

### Select the gages falling within a given county

Now let's zoom into a particular county and examine the gages located
there. 22. Select Lancaster county from your projected county sf
dataframe 23. Select the gage sites falling `within` that county \* Use
either matrix subsetting or tidy filtering 24. Create a plot showing: \*
all Nebraska counties, \* the selected county, \* and the gage sites in
that county

```{r spatial selection}
#22 Select the county
lancaster_county <- subset(NEcounties_sf_utm, NAME == "Lancaster")

#23 Spatially select gages within the selected county
gages_in_lancaster <- st_intersection(gage_sf_utm, lancaster_county)


#24 Plot
mapview(NEcounties_sf_utm, zcol = NULL, col.regions = "lightblue") +  
  mapview(lancaster_county, color = "orange") +  
  mapview(gages_in_lancaster, color = "blue")  

```
