---
title: "Assignment 2: Coding Basics"
author: "Sydney Williams" 
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW
```{r}

```

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 30, increasing by threes. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.
```{r}

```

4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. #Creating a sequence of numbers from one to 30, increasing by threes. 
sequence_1 <- seq(1,30,3)
sequence_1 

#2. #Computing the mean and median of the sequence with the mean and median functions
mean <- mean(sequence_1)
median <- median(sequence_1)

#3. #Creating a function to ask R whether mean is greather than median 
mean > median 

```

## Basics, Part 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

9.  QUESTION: How is this data frame different from a matrix?

> Answer: A data frame can hold multiple types of data, while a matrix can only have one type of data. Data frames have column names, matrices do not, you refer to the data point using row and column indices. 

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.


```{r basics 3}
# Create vectors for names, test scores, and pass status
names_vector <- c("James", "Ben", "Jordan", "Betty")
scores_vector <- c(68, 82, 95, 45)  
pass_status_vector <- scores_vector >= 50 

# This is the character vector 
names_vector <- c("James", "Ben", "Jordan", "Betty")

# This is the numerical vector out of a total of 100 points 
scores_vector <- c(68, 82, 95, 45) 

# This is the logical vector that shows TRUE if passed and FALSE if failed
pass_status_vector <-scores_vector >= 50 

# Combine vectors into a data frame
students_test_data <- data.frame(
  Names = names_vector,
  TestScores = scores_vector,
  PassStatus = pass_status_vector)

# Labeling the columns with informative names
colnames(students_test_data) <- c("Student Name","Test Score (out of 100)", "Pass/Fail")
print(students_test_data)

# 11a.  Function for passing score 
passing_score <- function(test_scores) {
  result <- ifelse(test_scores >= 50, TRUE, FALSE)
  print(result)  }

# 11b. Function for vector data from Question 5 
passing_score(scores_vector)
```


12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: ifelse worked. "if and else" gave an error that said condition had a length >1. 
