---
title: "Assignment 7: GLMs (Linear Regressions, ANOVA, & t-tests)"
author: "Sydney Williams"
date: "Spring 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
#setting up working directory 
getwd()

#loading all libraries needed for this assignment 
library(tidyverse)
library(agricolae)
library(here)  
library(lubridate)

# Importing raw data for chemistry/physics
chemistry_physics_rawdata <- read_csv("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv")

# Converting date columns to objects
chemistry_physics_rawdata$sampledate <- mdy(chemistry_physics_rawdata$sampledate)

#2
# Set ggplot theme as default
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "pink"), 
        legend.position = "top")
theme_set(mytheme)
```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: There is no significant difference in mean lake temperature recorded during July across different depths in all lakes.
Ha: There is a significant difference in mean lake temperature recorded during July across different depths in all lakes.

4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4

# Filtering for only dates in July with specific columns and removing NAs
july_data_2 <- chemistry_physics_rawdata %>%
     mutate(month = month(sampledate)) %>%
     filter(month == "7") %>%
     select(lakename, year4,daynum, depth, temperature_C) %>%
     drop_na(c("depth", "temperature_C"))
    
#5
# creating scatter plot of temperature by depth to visualize the relationship between them 
# added a smoothed line to show linear model and limit temp values 
# added other aesthetics to make plot look pretty and easy to read 
july_scatter_plot <- ggplot(july_data_2, aes(x = depth, y = temperature_C)) +
  geom_point(aes(color = lakename)) +  
  geom_smooth(method = "lm") +  
  labs(x = "Depth", y = "Temperature (°C)",  # Axes labels
       title = "Relationship between Depth and Temperature") + 
  ylim(limits = c(0, 35))

july_scatter_plot

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: As the depth increases, the temperature decreases. The distribution of the points suggests that there is a linear trend which indicates a strong relationship between temperature and depth. 


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7
# Performing linear regression
july_data.regression<- lm(temperature_C ~ depth, data = july_data_2)

# Displaying the results
summary(july_data.regression)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: The linear regression model suggests that depth has a statistically significant effect on temperature (p < 2.2e-16). The R-squared value of 0.744 and 392 DF indicates that the variability in temperature can be explained by changes in depth. For every 1-meter increase in depth, the model predicts a change in temperature of  degrees Celsius.

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
# Run AIC to determine set of explanatory variables is best suited to predict temperature
temperature.model <- step(lm(data = july_data_2,temperature_C ~ year4 + daynum + depth))

summary(temperature.model)


#10
# Run multiple regression on recommended set of variables
best.regression <- lm(temperature_C ~ year4 + daynum, data = july_data_2)

summary(best.regression)



```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The final set of explanatory variables that the AIC method suggests are year4 and daynum  to predict temperature in our multiple regression. The observed variance based on the R-squared value is 0.03216. This R-squared value is less than the R-squared value from the model using only depth as the explanatory variable. 

---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12

# Running an ANOVA model
July.anova <- aov(temperature_C ~ lakename, data = july_data_2)

# Running linear model 
July.linear <- lm(temperature_C ~ lakename, data = july_data_2)

summary(July.anova)
summary(July.linear)


```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes, there is a significant difference in mean temperature among the lakes as the p value is 0.3571 > 0.05. 


14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
#Creating graph that depicts temperature by depth with separate color for each lake 
scatter_plot_2 <- ggplot(data = july_data_2, aes(x = depth, y = temperature_C, color = lakename)) +
  geom_point(alpha = 0.5) + 
  geom_smooth(method = "lm", se = FALSE) +  
  labs(x = "Depth", y = "Temperature (°C)", 
       title = "Temperature by Depth for Different Lakes in July") +  
  ylim(0, 35)  

# Display the plot
print(scatter_plot_2)
```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
tukey_test <- TukeyHSD(July.anova)

tukey_test
```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: The lakes that have the same mean temperature, statistically speaking as Peter Lake are Ward Lake, Paul Lake and East Long Lake. 


17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: We might explore the Mann-Whitney U test to see whether Peter Lake and Paul Lake have distinct mean temperatures. 

18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
# Filter July data for Crampton Lake and Ward Lake
july_data_crampton_ward <- july_data_2 %>%
  filter(lakename %in% c("Crampton Lake", "Ward Lake"))

t.test(july_data_crampton_ward$temperature_C ~ july_data_crampton_ward$lakename)

```

>Answer: The degrees of freedom is 200.37. The p-value is 0.2649. The p-value is more that 0.05 so the null hypothesis that their means are equal is not rejected or they are statistically the same. The means for both lakes are similar with Crampton Lake at 15.35189 and Ward Lake at 14.45862. This corresponds with the results of the Tukey HSD test where the p value is more than 0.05 and the difference in mean is not significant. 